---
title: "Regression Project – Group 3 (Intermediate Report)"
author: "Reema Abolahum and Muzammil Mohammed"
date: 12.15.2025
embed-resources: true
toc: true
format: 
  html: 
    code-link: true
    code-tools: true
    toc-location: right
    df-print: paged
  pdf: 
    number-sections: true
  revealjs:
    output-ext: "revealjs-PRESENTATION.html"
    toc: false
    code-line-numbers: false
    echo: true
    scrollable: true
    code-link: true
    code-tools: true
    df-print: paged
    slide-number: true
---


```{r}
suppressPackageStartupMessages({
  library(tidyverse)
  library(readxl)
  library(simFrame)
  library(dplyr)        
  library(ggplot2)      
  library(tidyr)        
  library(forcats)      
  library(effects)  
  library(gt)
  library(corrplot)
  library(splines)
  library(car)
  library(ggeffects)
  library(GGally)
  library(effects)

})
```




# 1. Introduction

Employment income is a central indicator of economic well being, yet it is shaped by a complex interplay of demographic and social factors. In this report, we investigate how gender, citizenship, age, and household size influence income levels in South Austria, focusing specifically on the regions of Carinthia and Styria using the EU SILC panel dataset.


Our analysis proceeds in three stage:

•	Data preparation and descriptive statistics to identify distributions, outliers, and basic associations.

•	Regression modeling using linear, polynomial, and spline specifications to capture both straightforward and nonlinear relationships.

•	Interaction effects to explore how demographic variables combine to produce differentiated income outcomes.

From a methodological perspective, we employ:

•	Polynomial regression to model quadratic curvature in age and household size effects.

•	Spline regression to flexibly capture turning points and plateaus in continuous predictors.

•	Interaction terms to test whether the impact of one variable (e.g., age) depends on another (e.g., gender or citizenship).

•	Diagnostic checks to validate model assumptions and ensure robustness.

By combining descriptive visualization with advanced regression techniques, this report provides a nuanced view of income determinants in South Austria. The findings highlight both individual and joint effects of demographic factors, offering insights into inequality patterns and the importance of flexible modeling approaches.


# 2. Data Collection and Preparation
```{r}
data("eusilcP")

dat <- eusilcP %>%
  select(py010n, gender, citizenship, hsize, age, region) %>%
  filter(region %in% c("Carinthia", "Styria")) %>%
  filter(py010n > 0)
dat <- dat %>% rename(income = py010n)
dat$hsize <- as.numeric(as.character(dat$hsize))
dat <- na.omit(dat)

```

# 3. Descriptive Analysis
## 3.1 Numeric Summaries
```{r}
num_vars <- dat %>% select(income, age, hsize)
summary(num_vars)
```

## 3.2 Frequency Tables
```{r}
table(dat$gender)
table(dat$citizenship)
table(dat$gender, dat$citizenship)
```

## 3.3 Univariate Plots
```{r}
# Income
ggplot(dat, aes(y = income)) + 
  geom_boxplot() + 
  labs(title = "Income Boxplot")

ggplot(dat, aes(x = income)) + 
  geom_histogram(bins = 50) + 
  labs(title = "Income Histogram")

# Age
ggplot(dat, aes(y = age)) + 
  geom_boxplot() + 
  labs(title = "Age Boxplot")

ggplot(dat, aes(x = age)) + 
  geom_histogram(bins = 30) + 
  labs(title = "Age Histogram")

# Household Size
ggplot(dat, aes(y = hsize)) + 
  geom_boxplot() + 
  labs(title = "Household Size Boxplot")

ggplot(dat, aes(x = hsize)) + 
  geom_histogram(binwidth = 1) + 
  labs(title = "Household Size Histogram")

# Gender
ggplot(dat, aes(x = gender)) + 
  geom_bar() + 
  labs(title = "Gender Frequency")

# Citizenship
ggplot(dat, aes(x = citizenship)) + 
  geom_bar() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) + 
  labs(title = "Citizenship Frequency")

```

## 3.4 Bivariate Plots
```{r}
# Gender vs Income
ggplot(dat, aes(x = gender, y = income)) +
  geom_boxplot() +
  labs(title = "Income vs Gender", y = "Income (€)", x = "Gender") +
  theme_minimal()

# Citizenship vs Income
ggplot(dat, aes(x = citizenship, y = income)) +
  geom_boxplot() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Income vs Citizenship", y = "Income (€)", x = "Citizenship") +
  theme_minimal()

# Age vs Income (Polynomial)
ggplot(dat, aes(x = age, y = income)) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = "lm", formula = y ~ poly(x, 2), color = "blue") +
  labs(title = "Income vs Age (2nd-degree polynomial)",
       x = "Age (years)", y = "Income (€)") +
  theme_minimal()

# Household size vs Income (Polynomial)
ggplot(dat, aes(x = hsize, y = income)) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = "lm", formula = y ~ poly(x, 2), color = "blue") +
  labs(title = "Income vs Household Size (2nd-degree polynomial)",
       x = "Household Size", y = "Income (€)") +
  theme_minimal()

```

## 3.5 Interaction Plots
```{r}
# Gender × Citizenship
ggplot(dat, aes(x = gender, y = income, fill = citizenship)) +
  geom_boxplot(position = "dodge") +
  labs(title = "Income Interaction: Gender × Citizenship",
       y = "Income (€)", x = "Gender") +
  theme_minimal()

# Age × Gender
ggplot(dat, aes(x = age, y = income, color = gender)) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = "lm") +
  labs(title = "Income Interaction: Age × Gender",
       x = "Age (years)", y = "Income (€)") +
  theme_minimal()

# Age × Citizenship
ggplot(dat, aes(x = age, y = income, color = citizenship)) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = "lm") +
  labs(title = "Income Interaction: Age × Citizenship",
       x = "Age (years)", y = "Income (€)") +
  theme_minimal()

# Household Size × Gender
ggplot(dat, aes(x = hsize, y = income, color = gender)) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = "lm") +
  labs(title = "Income Interaction: Household Size × Gender",
       x = "Household Size", y = "Income (€)") +
  theme_minimal()

# Household Size × Citizenship
ggplot(dat, aes(x = hsize, y = income, color = citizenship)) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = "lm") +
  labs(title = "Income Interaction: Household Size × Citizenship",
       x = "Household Size", y = "Income (€)") +
  theme_minimal()

```

## 3.6 Correlation / Scatterplot Matrix
```{r}


# Correlation matrix
cor_mat <- cor(num_vars)
corrplot(cor_mat, method = "ellipse", tl.col = "black", tl.srt = 45)

# Scatterplot matrix (modern version)
ggpairs(num_vars,
        title = "Scatterplot Matrix of Income, Age, Household Size",
        upper = list(continuous = "cor"),
        lower = list(continuous = "smooth"),
        # Option A: histograms with custom binwidth
        diag = list(continuous = wrap("barDiag", binwidth = 5)))


```

# 4. Regression Modeling
## 4.1 Linear Model
```{r}

# Define linear model with renamed variable
lm_model <- lm(income ~ gender + citizenship + hsize + age, data = dat)

# Gender effect
plot(Effect("gender", lm_model),
     main = "Linear Effect: Gender")

# Citizenship effect
plot(Effect("citizenship", lm_model),
     main = "Linear Effect: Citizenship")

# Age effect (linear)
plot(Effect("age", lm_model,
            xlevels = list(age = seq(20, 80, by = 5))),
     main = "Linear Effect: Age")

# Household size effect (linear)
plot(Effect("hsize", lm_model,
            xlevels = list(hsize = 1:9)),
     main = "Linear Effect: Household Size")

```
## 4.2 Polynomial Model
```{r}
# Define polynomial model with renamed variable
poly_model <- lm(income ~ poly(age, 2) + poly(hsize, 2) + gender + citizenship,
                 data = dat)

# Gender effect
plot(Effect("gender", poly_model),
     main = "Polynomial Effect: Gender", ylab = "Predicted Income (€)")

# Citizenship effect
plot(Effect("citizenship", poly_model),
     main = "Polynomial Effect: Citizenship", ylab = "Predicted Income (€)")

# Age effect (quadratic)
plot(Effect("age", poly_model,
            xlevels = list(age = seq(20, 80, by = 5))),
     main = "Polynomial Effect: Age", ylab = "Predicted Income (€)")

# Household size effect (quadratic)
plot(Effect("hsize", poly_model,
            xlevels = list(hsize = 1:9)),
     main = "Polynomial Effect: Household Size", ylab = "Predicted Income (€)")

```
## 4.3 Spline Model
```{r}

# Define spline model with renamed variable

spline_model <- lm(income ~ ns(age, df = 4) + ns(hsize, df = 4) + gender + citizenship,
                   data = dat)

# Gender effect
plot(Effect("gender", spline_model),
     main = "Spline Effect: Gender", ylab = "Predicted Income (€)")

# Citizenship effect
plot(Effect("citizenship", spline_model),
     main = "Spline Effect: Citizenship", ylab = "Predicted Income (€)")

# Age effect (spline)
plot(Effect("age", spline_model,
            xlevels = list(age = seq(20, 80, by = 5))),
     main = "Spline Effect: Age", ylab = "Predicted Income (€)")

# Household size effect (spline)
plot(Effect("hsize", spline_model,
            xlevels = list(hsize = 1:9)),
     main = "Spline Effect: Household Size", ylab = "Predicted Income (€)")


```



## 4.4 Interaction Effects

```{r}

# Define interaction model with renamed variable

int_model <- lm(income ~ (gender + citizenship + ns(age, 3) + ns(hsize, 3))^2,
                data = dat)

# Age × Household Size interaction using ggeffects
eff_age_hsize <- ggpredict(int_model,
                           terms = c("age [20:80 by=5]", "hsize [1:9]"))

ggplot(eff_age_hsize, aes(x, predicted, color = group)) +
  geom_line(size = 1) +
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high, fill = group),
              alpha = 0.2, color = NA) +
  labs(title = "Interaction: Age × Household Size",
       x = "Age (years)", y = "Predicted Income (€)") +
  theme_minimal() +
  theme(legend.title = element_blank())


```

## 4.5 Interaction Plots

```{r}
# 4.5 Interaction Plots using effects 

int_model <- lm(income ~ (gender + citizenship + ns(age, 3) + ns(hsize, 3))^2,
                data = dat)

# Interaction plots
plot(Effect(c("gender", "citizenship"), int_model),
     main = "Interaction: Gender × Citizenship", ylab = "Predicted Income (€)")

plot(Effect(c("gender", "age"), int_model,
            xlevels = list(age = seq(20, 80, by = 5))),
     main = "Interaction: Gender × Age", ylab = "Predicted Income (€)")

plot(Effect(c("gender", "hsize"), int_model,
            xlevels = list(hsize = 1:9)),
     main = "Interaction: Gender × Household Size", ylab = "Predicted Income (€)")

plot(Effect(c("citizenship", "age"), int_model,
            xlevels = list(age = seq(20, 80, by = 5))),
     main = "Interaction: Citizenship × Age", ylab = "Predicted Income (€)")

plot(Effect(c("citizenship", "hsize"), int_model,
            xlevels = list(hsize = 1:9)),
     main = "Interaction: Citizenship × Household Size", ylab = "Predicted Income (€)")





```


# 5. Summary
Income in South Austria is highly skewed, with clear differences by gender and citizenship. Age and household size show nonlinear effects, best captured by polynomial and spline models. Interaction analyses reveal that demographic factors combine in complex ways to shape income outcomes. Overall, flexible regression approaches provide a more accurate picture than simple linear models, and diagnostics confirm model adequacy